\bf Model Visualization Techniques for a Social Network Model

### 2017-08-17 paper.Rnw
pages 1-5

\section{Introduction} 
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }
\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
\st{XXX - okay, here is what i have done for that:
Social networks have been studied for decades, beginning with a few foundational works, the most well known of which is the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations. In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data. There are many kinds of social networks, but there are not as many statistical models for social network data.} Some network models that have been applied to social networks include the exponential random graph model and latent space models. These models, however, are only for single instance networks. If we only care about one network observation, that is not a problem, but if we have many observations of social network over time, these models may not be appropriate. When studying a network over time, which is referred to as a \textit{dynamic network}, we need a model that can take time into account. 
Models for dynamic social networks are extremely important because of how realistic they are. Social networks do not form spontaneously; they evolve over time. Ties are added and removed, and new actors join the network. Modeling the underlying mechanisms that create network changes over time is very complex but also has the potential to uncover hidden truths.

\par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, when other models only rely on  the network statistics to determine change. Letting the actor-level statistics change the structure of the network leads to a more practical and relevant approach to model change in a social network: in the ``real" world we expect people with common interests to be more likely to form relationships. SAOMs allow us to incorporate this intuition in the modeling process.

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. \st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically. Likelihood functions turn into very complex objects due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, 
SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators.} Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs, we help researchers working with the models better understand the implications and analysis of these models.
}

\par The guiding principles of model visualization introduced in \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore the fitting process rather than just looking at the final result. Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization. First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process. Bringing the CTMC out of the black box and into the light of visualization  provides researchers with insights into the underlying features of the model. Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, and so are the associated parameters. By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects. Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks for these estimates rely on simulations from the fitted model. Again, these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring these elements and others into the foreground, leading to a better understanding and higher accessibility of stochastic actor-oriented models  for social network analysts.

In Section~\ref{sec:netintro}, we introduce the concepts of networks and network visualizations. In Section~\ref{sec:saoms}, we present the SAOM for social network analysis. In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.

@@

Network data is of frequent interest to researchers in a wide array of fields. There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, e.g. neural networks, and social networks, just to name a few \citep{newman}. Each of these examples have one thing in common: their data structure. There are always units of observation, which we refer to throughout this paper as \textit{nodes}, and connections of some kind between those units, which we will call \textit{edges}. Networks might change over time, for example when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network. Nodes and edges themselves also have inherent variables of interest, e.g.\ the affiliation of authors in a co-authorship network, or the number of times two authors have collaborated.

The multiple layers of network data structures create unique struggles to network analysts. Some questions that researchers aim to answer are: How does the strength of a connection between two nodes affect the overall structure of the network? Do node-level differences affect the formation or destruction of edges? Which views of the data are most informative for communicating effects and results of statistical analyses? These are, of course, just a few broad questions. In this paper we want to focus on the visual exploration of network data and models.

@@

There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \citep{knuth2000, adjmatpro}. As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.

@@

The first method, the node-link diagram, represents nodes with markers in 2D Euclidean space and then represents edges by connecting the markers with lines when there is an edge between the two nodes. But because there is no natural placement of the points, a random placement is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.

Some layout algorithms were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. The network's edges act as springs pushing and pulling the nodes in a lower dimensional (usually two-dimensional) space. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes. 
\hh{XXX how does the adjacency matrix feature here? The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below. XXX }
Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, treat networks like physical systems. Other algorithms are based on multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix \citep{drawingalgs}. The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.

@@

<<sienagof, fig.cap="Goodness of fit measures for two network statistics: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data. \\st{The p-value shown on the x-axis is .... }", cache = TRUE, message = FALSE, warning = FALSE>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship")
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship")
p1 <- plot(gof1)
p2 <- plot(gof2)
gridExtra::grid.arrange(p1, p2)
@

>>>>

\section{Introduction} 
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }
\hh{XXX - talk a bit about social networks first, why we want to model them, and that usually the models are not dynamic.}
\st{XXX - okay, here is what i have done for that:
Social networks have been studied for decades, beginning with a few foundational works, including the 1967 study, "The Small World Problem" by Stanley Milgram \citep{goldenberg09}. Examples of social networks include collaboration networks between academic researchers, friendship networks in a school or university, and trade networks between nations. In recent years, the study of social networks has grown in popularity due to an increase in the availability and access to social network data. There are many kinds of social networks, but there are not as many statistical models for social network data.} Some network models that have been applied to social networks include the exponential random graph model and latent space models. These models, however, are only for single instance networks. If we only have one network observation, or only care about one state of a complex network in time, like a snapshot of the World Wide Web, using the well-established models for single network observating is not a problem. If, on the other hand, we have many observations of social network over time, these models may not be appropriate because they do not explicitly allow for the network to change as time passes. When studying a network over time, referred to as a \textit{dynamic network}, we need a model that can take the time aspect of the network into account. 
Models for dynamic social networks have a great deal of modelling potential because of how realistic their structure can be. A social network does not form spontaneously: it evolves over time. Ties are formed and dissolved, and new actors join the social structure. Modeling the underlying mechanisms that create network changes over time is very complex but also provides potential to uncover hidden truths.

\par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network. Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network. In the ``real" world we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. \st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically. Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, 
SAOMs are often fit to data using a series of Markov chain Monte Carlo (MCMC) phases for finding method of moments estimators. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are ``behind the scenes" of when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
}

\par The guiding principles of model visualization introduced in \citet{modelvis} are: 
\begin{enumerate}
\item Display the model in the data space.
\item Collections of models are more informative than singletons. 
\item Explore the fitting \textit{process} rather than just looking at the final result.
\end{enumerate}
Stochastic actor-oriented models are a prime example of a set of models that can benefit greatly from application of model visualization. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Bringing the CTMC out of the black box and into the light through model visualization can provide researchers with insights into the underlying features of the model. Furthermore, SAOMs can include a great deal of parameters to be added to the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects in the model. In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model. Again, each of these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

In Section~\ref{sec:netintro}, we introduce basic concepts of networks and network visualizations. In Section~\ref{sec:saoms}, we present the family of stochastic actor-oriented models for social network analysis. In Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.

@@

Network data is of frequent interest to researchers in a wide array of fields. There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, like neural networks, and social networks, just to name a few \citep{newman}. Each of these examples have one thing in common: their data \textit{structure}. There are always units of observation: the power stations, websites, neurons, and people, which we refer to throughout this paper as \textit{nodes} or \textit{actors}. There are also always connections of some kind between those units: the power lines, hyperlinks, electrical signals, and relationships, which we will call \textit{edges} or \textit{ties}. Networks might change over time, like when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network. The nodes and edges themselves can also have inherent variables of interest, e.g.\ the institution of authors in a co-authorship network, or the number of times two authors have collaborated.

The multiple layers of network data structures pose unique problems to network analysts. Some questions that network researchers may aim to answer are: How does the strength of a tie between two nodes affect the overall structure of the network? Do node-level differences affect the formation or dissolution of edges? Which views of the data are most informative for communicating significant effects and other results of statistical analyses of the network of interest? These are, of course, just a few broad questions, and we focus here on the latter, which we aim to answer through visual exploration of network data and models.

@@

There are two primary methods used to visualize networks: node-link diagrams and adjacency matrix visualization \citep{knuth2000, adjmatpro}. As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.

@@

The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes. These lines can also have arrows on them indicating the direction of the edge for directed networks. But because there is typically no natural placement of the points unless they have important spatial locations, a random placement of the points is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.

Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space. Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes. Yet another layout algorithm relies on the eigenstructure of the network's adjacency matrix to place the nodes in 2D space \citep{drawingalgs}. The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes. 
\hh{XXX how does the adjacency matrix feature here? The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below. XXX }

@@

<<sienagof, fig.cap="Goodness of fit measures for two network statistics: indegree and outdegree cumulative distributions. On the x-axis is the degree value, and on the y-axis is the number of times that degree value occurs in the data. \\st{The p-value shown on the x-axis is .... }", cache = TRUE, message = FALSE, warning = FALSE>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
gof2 <- sienaGOF(ansnull, IndegreeDistribution, varName ="friendship", join = FALSE)
p1 <- plot(gof1, period = 1)
p2 <- plot(gof2, period = 1)
gridExtra::grid.arrange(p1, p2)
# sum(gof1$`Period 1`$SimulatedTestStat >= gof1$`Period 1`$ObservedTestStat)/1000
# [1] 0.154
# > sum(gof2$`Period 1`$SimulatedTestStat >= gof2$`Period 1`$ObservedTestStat)/1000
# [1] 0.063

@
